---
title: "Fitting a TOPALS mortality model with age-grouped data, by Penalized Iteratively Weighted Least Squares (PIRLS)"
author: "Carl Schmertmann"
date: "21 Aug 2019, updated 14 Oct 2021 (cleaned up two minor notational errors)"
output:
  pdf_document: default
  html_document: default
header-includes: 
  - \usepackage{setspace}
  - \onehalfspacing
fontsize: 11pt
indent: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Main Idea

The objective is to fit a TOPALS model mortality schedule for $A$ single years of age $x=0,1,\ldots,(A-1)$ from exposure and death data for $G\le A$ age groups: $N_1\ldots N_G$ and deaths $D_1\ldots D_G$.

## Sample Data

Observed data consists of deaths $D_g$ and exposure $N_g$ for closed, non-overlapping age groups $g=1\ldots G$. Age groups are defined by a vector of $G+1$ boundary ages -- e.g. if the boundary ages are $0,1,5,10,\ldots,85,90$ then the age groups are $[0,1), [1,5),\ldots,[85,90)$. Denote $X_g$ as the set of integer ages that belong to group $g$, and $n_g$ as the number of ages in $X_g$.  

Typical age groups have boundaries like $0,1,5,10,\ldots,85,90$. However, if single-year death and exposure data is available then the "groups" could be simply integer ages, in which case boundaries for $A$ single-year "groups" would be $0,1,2,\ldots,A$ 

## Poisson Log Likelihood for Grouped Data

Regardless of the age grouping in the data, we assume that there is a latent schedule for $A$ single-year ages. Deaths at each single-year age have a Poisson distribution with expected value equal to the (possibly unobserved) exposure $N_x$ times the mortality rate: 
$$
D_x \sim Poisson \left( N_x\,\mu_x \right)\quad x=0,1,2,\ldots (A-1)
$$
Under the standard assumption that age-specific deaths are statistically independent, this implies that deaths  are also Poisson distributed within each age group:
$$
D_g \sim Poisson \left( \sum_{x \in X_g} N_x \, \mu_x \right) \quad g=1,2,\ldots G
$$
Expressing this in terms of the total _observed_ exposure in each age group,
$$
D_g \sim Poisson \left( N_g \sum_{x \in X_g} \left[ \frac{N_x}{N_g} \right] \mu_x \right)
$$
or 
$$
D_g \sim Poisson \left( N_g\, M_g \right)
$$
where $M_g$ represents the exposure-weighted average mortality rate across ages in group $g$. 

In the absence of single-year exposure data, we approximate with $M_g = [\tfrac{1}{n_g}]\sum_{x \in X_g} \mu_x$.\footnote{
An alternative approach might interpolate single-year exposure $(N_0,N_1,\ldots N_{A-1})$ from the available $N_g$ to construct more nuanced weights for intra-group averaging of mortality rates. In most populations this would have only minor effects on estimates. 
}

\noindent The entire vector of $G$ averaged mortality rates by age group is therefore
$$
\boldsymbol{M} = 
\begin{pmatrix}
n_1^{-1}\ldots n_1^{-1} & \cdots & 0\ldots0 \\
\vdots  & \ddots & \vdots  \\
0\ldots0 & \cdots & n_G^{-1}\ldots n_G^{-1}
\end{pmatrix}
\begin{pmatrix}
\mu_0 \\
\vdots \\
\mu_{A-1} \\
\end{pmatrix} 
\;=\; \boldsymbol{W} \boldsymbol{\mu} 
$$
where $\boldsymbol{W}$ is a $G \times A$ matrix of weights with row sums all equal to one.\footnote{
When "groups" correspond to single years, $\boldsymbol{W}=\boldsymbol{I}_A$ and $\boldsymbol{M}=\boldsymbol{\mu}$.
} 

The sample log likelihood for a single-year mortality schedule $\boldsymbol{\mu}$ is thus
$$
\begin{aligned}
  \ln L(\boldsymbol{\mu}) &= c + \sum_{g=1}^{G} \left(\, D_g \ln M_g - N_g M_g\,\right) \\
   &= c + \sum_{g=1}^{G} \left(\, D_g \ln M_g - \hat{D}_g \right) 
\end{aligned}  
$$
where $\hat{D}_g=N_g M_g$ represents the expected number of deaths in group $g$. It will also be useful to express the likelihood in matrix terms:

\begin{equation}
\label{eq:vec-lik}
\ln L = 
c + \left[ \ln M_1 \cdots \ln M_G \right] \boldsymbol{D}_{obs} -  \hat{\boldsymbol{D}}^\prime \boldsymbol{1} 
\end{equation}

\noindent where $\boldsymbol{D}_{obs}$ and $\hat{\boldsymbol{D}}$ are $G\times 1$ vectors of observed and expected deaths, ordered by age group, and $\boldsymbol{1}$ is a $G\times1$ vector of ones.

## TOPALS Model Mortality Schedule

The TOPALS model for single-year mortality rates is
$$
\boldsymbol{\lambda} = \boldsymbol{\lambda}^{\ast} + \boldsymbol{B} \boldsymbol{\alpha}
$$
where $\boldsymbol{\lambda}$ is an $A \times 1$ vector of age-specific log mortality rates for ages $0\ldots (A-1)$, $\boldsymbol{\lambda}^{\ast}$ s a vector with fixed constants representing a standard schedule for those ages, $\boldsymbol{B}$ is a $A \times K$ matrix of linear B-spline constants\footnote{Typically $K=7$, with spline knots fixed at ages $t=0,1,10,20,40,70,99$. Values in the $k$th column of $B$ are 
$$
  B_{xk} = 
\begin{cases}
\frac{x-t_{k-1}}{t_k-t_{k-1}}  &\mbox{for  }  x \in  [t_{k-1},t_k] \\
\frac{t_{k+1}-x}{t_{k+1}-t_k} &\mbox{for  }  x \in  [t_k,t_{k+1}] \\
0                              &\mbox{otherwise} 
\end{cases}
$$
}
, and $\boldsymbol{\alpha}$ is a $K$-dimensional vector representing deviations from the standard log mortality schedule at specified ages. 

In this model the log mortality rate at age $x \in \{0\ldots(A-1)\}$ is 
$$
\ln \mu_x = \lambda_x =  \lambda_x^{\ast} + \boldsymbol{b}_x^\prime \boldsymbol{\alpha}
$$
and the mortality rate is 
$$
\mu_x = exp \left( \lambda_x^{\ast} + \boldsymbol{b}_x^\prime \boldsymbol{\alpha} \right)
$$
where $\boldsymbol{b}_x^\prime$ is the $1 \times K$ row of $\boldsymbol{B}$ that corresponds to age $x$. 

### TOPALS derivatives 

Derivatives of age-specific rates with respect to TOPALS parameters $\boldsymbol{\alpha}$ are $K\times 1$ vectors: 
$$
\frac{\partial \ln \mu_x}{\partial \boldsymbol{\alpha}} = \,\boldsymbol{b}_x
\quad , \quad
\frac{\partial \mu_x}{\partial \boldsymbol{\alpha}} = \mu_x \,\boldsymbol{b}_x 
$$

\noindent For the entire $G \times 1$ vector of group mortality rates, this implies
$$
\begin{aligned}
\frac{\partial \boldsymbol{M}^\prime}{\partial \boldsymbol{\alpha}} &= 
\frac{\partial }{\partial \boldsymbol{\alpha}}\,(\boldsymbol{\mu}^\prime \boldsymbol{W}^\prime) \\& = 
\begin{pmatrix}
\frac{\partial \mu_0}{\partial \boldsymbol{\alpha}}
& \cdots  
& \frac{\partial \mu_{A-1}}{\partial \boldsymbol{\alpha}}
\end{pmatrix} \boldsymbol{W}^\prime \\
&=
\begin{pmatrix} 
\mu_0 \, \boldsymbol{b}_0 
& \cdots  
& \mu_{A-1} \, \boldsymbol{b}_{A-1}
\end{pmatrix} \boldsymbol{W}^\prime \\
&= \boldsymbol{B}^\prime \, diag(\boldsymbol{\mu}) \,\boldsymbol{W}^\prime
\end{aligned}
$$
\noindent Abbreviate this as 
$$
\frac{\partial \boldsymbol{M}^\prime}{\partial \boldsymbol{\alpha}} = \boldsymbol{X}^\prime
$$
\noindent remembering that the $G \times K$ matrix $\boldsymbol{X}=\boldsymbol{W} \, diag(\boldsymbol{\mu}) \,\boldsymbol{B}$ varies with parameters $\boldsymbol{\alpha}$ via the $\mu$ terms in the central diagonal matrix.


The $G \times 1$ vector of expected deaths in groups $1\ldots G$ is
$$
\hat{\boldsymbol{D}} = \begin{pmatrix}
N_1 M_1 \\
\vdots \\
N_G M_G \\
\end{pmatrix} = diag(\boldsymbol{N})\, \boldsymbol{M}
$$
and the 
derivative of its transpose with respect to TOPALS parameters $\boldsymbol{\alpha}$ is a $K \times G$ matrix:

\begin{equation}
\label{eq:dg-dalpha}
\frac{\partial \hat{\boldsymbol{D}}^\prime}  
{\partial \boldsymbol{\alpha}} =   
\frac{\partial \boldsymbol{M}^\prime}{\partial \boldsymbol{\alpha}}\, diag(\boldsymbol{N})\,
=
 \boldsymbol{X}^\prime diag(\boldsymbol{N})
\end{equation}


### Penalized Log Likelihood for TOPALS parameters

With the TOPALS parameterization the log likelihood for a sample $\{D_g,N_g\}$ is
$$
\begin{aligned}
  \ln L(\boldsymbol{\alpha}) 
&=  c + \sum_{g=1}^{G} \left(\, D_g \ln M_g(\boldsymbol{\alpha}) - \hat{D}_g(\boldsymbol{\alpha}) \right) \\
&=  c + \left[ \ln M_1 \cdots \ln M_G \right] \boldsymbol{D}_{obs} -  \hat{\boldsymbol{D}}^\prime \boldsymbol{1} 
 \end{aligned}
$$

\noindent To stabilize estimates in small populations with few deaths, we add a small penalty term to the log likehood:

\begin{align}
\begin{split}
\label{eq:Qa-defn}
  Q(\boldsymbol{\alpha}) 
  &= \ln L(\boldsymbol{\alpha}) - \text{penalty}(\boldsymbol{\alpha}) \\
  &= c + \left[ \ln M_1 \cdots \ln M_G \right] \boldsymbol{D} -  \hat{\boldsymbol{D}}^\prime \boldsymbol{1}  - \tfrac{1}{2}\,\boldsymbol{\alpha}^\prime \boldsymbol{P} \boldsymbol{\alpha}
\end{split}
\end{align}  

\noindent where $\boldsymbol{P}$ is a $K \times K$ matrix of constants selected so that the penalty term equals the sum of squared differences between consecutive  $\alpha$ parameters -- i.e.,  $\tfrac{1}{2}\boldsymbol{\alpha}^\prime \boldsymbol{P} \boldsymbol{\alpha}=\sum_{k=2}^K (\alpha_k-\alpha_{k-1})^2$. Adding the penalty gives priority to sets of TOPALS parameters $(\alpha_1\ldots \alpha_K)$ that are similar to one another, and thus to log mortality schedules that look more like up-and-down vertical shifts of the standard schedule. For all but the smallest populations the penalty term has virtually no effect on parameter estimates. For very small populations with zero deaths in some age groups, the addition of the penalty  stabilizes estimated schedules by borrowing strength across groups. 

### Maximizing via Iteratively Reweighted Regression

We want to select $\boldsymbol{\alpha}$ to maximize $Q$. This requires setting a vector of derivatives equal to zero: $\frac{\partial Q}{\partial\boldsymbol{\alpha}} = 0 \in \mathbb{R}^K$. 


Differentiating Eq. (\ref{eq:Qa-defn}) with respect to the TOPALS parameters and substituting some of the results above produces
$$
\begin{aligned}
  \frac{\partial Q}{\partial\boldsymbol{\alpha}} 
  &= 
   \left[ \tfrac{1}{M_1}\tfrac{\partial M_1}{\partial \boldsymbol{\alpha}}  \cdots
   \tfrac{1}{M_G}\tfrac{\partial M_G}{\partial \boldsymbol{\alpha}}  \right] \boldsymbol{D}_{obs} -  
\frac{\partial \hat{\boldsymbol{D}}^\prime}{\partial \boldsymbol{\alpha}} \boldsymbol{1}  -  \boldsymbol{P} \boldsymbol{\alpha}
 \\
&= \frac{\partial \boldsymbol{M}^\prime}{\partial \boldsymbol{\alpha}} diag(\frac{1}{\boldsymbol{M}}) \, \boldsymbol{D}_{obs} - \frac{\partial \boldsymbol{M}^\prime}{\partial \boldsymbol{\alpha}} \, diag(\boldsymbol{N}) \, \boldsymbol{1}
-  \boldsymbol{P} \boldsymbol{\alpha} \\
&= \boldsymbol{X}^\prime \, diag(\frac{1}{\boldsymbol{M}}) \, \boldsymbol{D}_{obs} - \boldsymbol{X}^\prime \boldsymbol{N} 
-  \boldsymbol{P} \boldsymbol{\alpha} \\
&= \boldsymbol{X}^\prime \, diag(\frac{1}{\boldsymbol{M}}) \, \boldsymbol{D}_{obs} - \boldsymbol{X}^\prime \, diag(\frac{1}{\boldsymbol{M}}) \hat{\boldsymbol{D}} 
-  \boldsymbol{P} \boldsymbol{\alpha} \\
&=  \boldsymbol{X}^\prime \, diag(\frac{1}{\boldsymbol{M}}) (\boldsymbol{D}_{obs} - \hat{\boldsymbol{D}}) 
- \boldsymbol{P}\boldsymbol{\alpha} 
\end{aligned}  
$$

\noindent For reasons that will become clear in a moment, it is convenient to define the abbreviation $\boldsymbol{A}=diag(\tfrac{N_1}{M_1}\ldots\tfrac{N_G}{M_G})$. Using this new abbreviation,

\begin{equation}
\label{eq:dQ-dalpha}
  \frac{\partial Q}{\partial\boldsymbol{\alpha}} =  \boldsymbol{X}^\prime 
\boldsymbol{A}\:  
  diag(\frac{1}{\boldsymbol{N}}) (\boldsymbol{D}_{obs} - \hat{\boldsymbol{D}})  
- \boldsymbol{P}\boldsymbol{\alpha} = \boldsymbol{0} \in \mathbb{R}^K
\end{equation}


Rearranging Eq. (\ref{eq:dQ-dalpha}) and placing hats over the  $\boldsymbol{\alpha}$-dependent terms yields a system of $K$ equations in $K$ variables $\boldsymbol{\alpha}$

\begin{equation}
\label{eq:max-conditions}
\hat{\boldsymbol{X}}^\prime \hat{\boldsymbol{A}} \: diag(\frac{\hat{\boldsymbol{D}}}{\boldsymbol{N}})  
+ \boldsymbol{P}\hat{\boldsymbol{\alpha}} = 
\hat{\boldsymbol{X}}^\prime \hat{\boldsymbol{A}} \: diag(\frac{\boldsymbol{D}_{obs}}{\boldsymbol{N}}) 
\end{equation}

The iteratively reweighted least squares approach to solving the system of equations in (\ref{eq:max-conditions}) begins by approximating how predicted values $\hat{\boldsymbol{D}}$ would change if the vector of TOPALS parameters changed from $\hat{\boldsymbol{\alpha}}$ to some other value $\boldsymbol{\alpha}^\ast$, while (temporarily) keeping matrices $\hat{\boldsymbol{X}}$ and $\hat{\boldsymbol{A}}$ constant:

\begin{equation}
\label{eq:start-IRLS}
\hat{\boldsymbol{X}}^\prime \hat{\boldsymbol{A}} \: diag(\frac{1}{\boldsymbol{N}}) \left[ \hat{\boldsymbol{D}} + \frac{\partial \hat{\boldsymbol{D}}}{\partial \boldsymbol{\alpha}^\prime} (\boldsymbol{\alpha}^\ast - \hat{\boldsymbol{\alpha}}) \right] 
+ \boldsymbol{P}\,\boldsymbol{\alpha}^\ast = 
\hat{\boldsymbol{X}}^\prime \hat{\boldsymbol{A}} \: diag(\frac{\boldsymbol{D}_{obs}}{\boldsymbol{N}}) 
\end{equation}

\noindent which from Eq. (\ref{eq:dg-dalpha}) expands to

\begin{equation}
\label{eq:second-IRLS}
\hat{\boldsymbol{X}}^\prime \hat{\boldsymbol{A}} \: diag(\frac{1}{\boldsymbol{N}}) \left[ \hat{\boldsymbol{D}} + 
diag(\boldsymbol{N}) \hat{\boldsymbol{X}} \:  \,
(\boldsymbol{\alpha}^\ast - \hat{\boldsymbol{\alpha}}) \right] 
+ \boldsymbol{P}\hat{\boldsymbol{\alpha}} = 
\hat{\boldsymbol{X}}^\prime \hat{\boldsymbol{A}} \: diag(\frac{1}{\boldsymbol{N}}) \boldsymbol{D}_{obs}
\end{equation}

\noindent This can be rearranged as 
$$
\hat{\boldsymbol{X}}^\prime \hat{\boldsymbol{A}} \:  
 \hat{\boldsymbol{X}} \:  \,
(\boldsymbol{\alpha}^\ast - \hat{\boldsymbol{\alpha}}) 
+ \boldsymbol{P} \boldsymbol{\alpha}^\ast = 
\hat{\boldsymbol{X}}^\prime \hat{\boldsymbol{A}} \: diag(\frac{1}{\boldsymbol{N}}) (\boldsymbol{D}_{obs} - \hat{\boldsymbol{D}})
$$
and further simplified as 
$$
\left[ \hat{\boldsymbol{X}}^\prime \hat{\boldsymbol{A}} \:  
 \hat{\boldsymbol{X}} \:  \,
+ \boldsymbol{P} \right] \boldsymbol{\alpha}^\ast = 
\hat{\boldsymbol{X}}^\prime \hat{\boldsymbol{A}} \: \left[ diag(\frac{1}{\boldsymbol{N}}) (\boldsymbol{D}_{obs} - \hat{\boldsymbol{D}})  + \hat{\boldsymbol{X}} \hat{\boldsymbol{\alpha}} \right]
$$
and finally simplified to
\begin{equation}
\label{eq:final-IRLS}
\left[ \hat{\boldsymbol{X}}^\prime \hat{\boldsymbol{A}} \:  
 \hat{\boldsymbol{X}} \:  \,
+ \boldsymbol{P} \right] \boldsymbol{\alpha}^\ast = 
\hat{\boldsymbol{X}}^\prime \hat{\boldsymbol{A}} \: \hat{\boldsymbol{y}}
\end{equation}
where $\hat{\boldsymbol{y}} = diag(\frac{1}{\boldsymbol{N}}) (\boldsymbol{D}_{obs} - \hat{\boldsymbol{D}})  + \hat{\boldsymbol{X}} \hat{\boldsymbol{\alpha}}$. 

The solution $\boldsymbol{\alpha}^\ast$ to Eq. (\ref{eq:final-IRLS}) is also the coefficient estimate from a (penalized) weighted least squares regression of $\hat{\boldsymbol{y}}$ on $\hat{\boldsymbol{X}}$ with weights $\hat{\boldsymbol{A}}^{-1}$. 

Starting from an arbitrary vector $\hat{\boldsymbol{\alpha}}$, one can solve for the vector that maximizes penalized likelihood $Q(\boldsymbol{\alpha})$ by 

1. calculating the values of ($\hat{\boldsymbol{X}},\hat{\boldsymbol{A}},\hat{\boldsymbol{D}}$) at the current parameter value
2. solving eq. (\ref{eq:final-IRLS}) for $\boldsymbol{\alpha}^\ast$, and 
3. repeating until convergence. 

\noindent This is the algorithm used in the **TOPALS_fit** function in the next sections.  


# Example: A TOPALS fit from five-year age groups for Italian Females in 1980

Age groups for this exercise are 0, 1--4, 5--9,...,80--85.  

```{r message=FALSE}
library(tidyverse, quietly = TRUE)
library(splines, quietly = TRUE)
```

Display the code for the _TOPALS_fit()_ function 
```{r, code=readLines("TOPALS_fit.R")}
```

```{r}

# include the fitting function
source('TOPALS_fit.R', echo=FALSE, local = knitr::knit_global())


## Italian female 1980 HMD data for age groups


boundaries = c(0,1,seq(5,85,5))  # last group is [80,85)


N = c(312106.85, 1423566.3, 2105814.63, 2249555.41, 2230885.74, 1983157.8, 
1874479.58, 1989351.99, 1772133.73, 1896866.51, 1836597.36, 1834496.64, 
1811178.38, 1192763.85, 1498384.03, 1223810.9, 863725.92, 537720.77)

D = c(3889, 716, 587, 589, 791, 816, 832, 1257, 1651, 2721, 4310, 
6636, 10536, 11043, 23312, 34945, 44537, 50392)

names(N) = names(D) = head(boundaries,-1)
  
## standard schedule is (very arbitrarily!) the 1959 period 
## schedule for Canadian females at ages 0,1,...,99 
## these are the log mortality rates
std = c(-3.8933, -5.7776, -6.8474, -7.3298, -7.4519, -7.4408, -7.4807, 
-7.5845, -7.7219, -7.8628, -7.9771, -8.041, -8.0568, -8.0329, 
-7.9779, -7.9005, -7.8088, -7.7101, -7.6113, -7.5195, -7.4415, 
-7.3823, -7.3393, -7.308, -7.2837, -7.2619, -7.238, -7.2082, 
-7.1712, -7.1264, -7.0735, -7.0118, -6.9414, -6.8648, -6.7849, 
-6.7047, -6.6272, -6.5544, -6.4845, -6.4147, -6.3423, -6.2645, 
-6.1791, -6.0872, -5.9904, -5.8903, -5.7887, -5.6869, -5.586, 
-5.4866, -5.3895, -5.2953, -5.2049, -5.1186, -5.0347, -4.9513, 
-4.8664, -4.778, -4.6847, -4.5877, -4.4887, -4.3895, -4.2918, 
-4.1969, -4.1041, -4.0122, -3.9199, -3.8261, -3.7296, -3.6303, 
-3.5278, -3.4221, -3.3129, -3.2004, -3.0861, -2.9716, -2.8589, 
-2.7497, -2.6458, -2.5482, -2.4556, -2.3659, -2.2771, -2.187, 
-2.0942, -1.9991, -1.9028, -1.8062, -1.7105, -1.6164, -1.5242, 
-1.434, -1.3458, -1.2596, -1.1758, -1.0958, -1.0212, -0.9535, 
-0.8944, -0.8454)

## single-year log mortality rates from HMD
## these are the targets for TOPALS estimation
ITA_HMD_logmx = 
  c(-4.3852, -7.1185, -7.6009, -7.7517, -8.1117, -8.1456, -8.1456, 
-8.1456, -8.294, -8.2171, -8.4684, -8.294, -8.3349, -8.1456, 
-8.0789, -7.9866, -7.9866, -8.0164, -7.902, -7.824, -7.7753, 
-7.7753, -7.7753, -7.7753, -7.8753, -7.7063, -7.7063, -7.6628, 
-7.6417, -7.8753, -7.4876, -7.4354, -7.2644, -7.3233, -7.3385, 
-7.2644, -7.0021, -6.959, -6.959, -6.7855, -6.8216, -6.5713, 
-6.5225, -6.4956, -6.3539, -6.2712, -6.2196, -6.0035, -5.9835, 
-5.8569, -5.7992, -5.7169, -5.6694, -5.5315, -5.433, -5.3247, 
-5.2514, -5.1814, -5.0625, -4.9533, -4.8783, -4.7915, -4.6767, 
-4.5923, -4.4945, -4.3836, -4.2992, -4.1825, -4.0513, -3.9409, 
-3.8135, -3.6913, -3.5332, -3.4455, -3.2966, -3.2069, -3.0614, 
-2.9677, -2.8466, -2.7201, -2.5974, -2.4617, -2.3462, -2.2249, 
-2.1253, -1.9713, -1.8905, -1.7861, -1.6842, -1.5945, -1.4583, 
-1.3792, -1.297, -1.2087, -1.1393, -1.0245, -0.9444, -0.8681, 
-0.7958, -0.7276)



show = function(fit, hue='red') {
 
 df_grouped = data.frame(
                L = head( fit$age_group_bounds, -1),
                U = tail( fit$age_group_bounds, -1),
                N = fit$N,
                D = fit$D
               ) %>%
          mutate(logmx_obs = log(D/N))
              
 
 df_single  = data.frame(
                  age=  seq(fit$std) - .50,  # 0.5, 1.5, ...
                  std = fit$std,
                  logmx_true = ITA_HMD_logmx,
                  logmx_fit  = fit$logm
               )

this_plot =
  ggplot(data = df_single, aes(x=age,y=logmx_true)) +
      geom_line(aes(x=age,y=std), color='black', lwd=0.5) +
      geom_line(aes(x=age,y=logmx_fit), color=hue, lwd=3, alpha=.40) +
      geom_segment(data=df_grouped,aes(x=L,xend=U,
                                       y=logmx_obs,
                                       yend=logmx_obs),
                   color=hue,lwd=1.5, alpha=.90) +
         geom_point(size=0.60) +
      labs(x='Age',y='Log Mortality Rate',
           title='Italy Females 1980',
           subtitle = paste(sum(fit$D),'deaths to',round(sum(fit$N)),'women')) +
      scale_x_continuous(breaks=c(0,1,seq(5,100,5)),minor_breaks = NULL) +
      scale_y_continuous(limits=c(-10,0),breaks=seq(-10,0,2),minor_breaks = NULL) +
      theme_bw()

  print(this_plot)
} # show  



```
Now fit the data and show the result.
```{r}

fit = TOPALS_fit(N,D,std,
                 age_group_bounds = boundaries,
                 details=TRUE)

str(fit)

show(fit)

```



In this figure the dots are the true rates from the HMD, the horizonal bars are the $\log(N_g/D_g)$ values from the grouped data, the black curve is the standard, and the red curve is the TOPALS single-year fit from PIRLS. 

# Example: Very small populations with Italy 1980 rates

Repeat the exercise in the last section but with simulated deaths for a much smaller population subject to the same rates.  Suppose there were only 10000 Italian women and simulate deaths:

```{r}


for (i in 1:4) {  
  ## simulate a small population    
    target_pop = 10000
    smallN = N * target_pop/sum(N)
    smallD = rpois(length(smallN), smallN* D/N)
    
    this_fit = TOPALS_fit(smallN, smallD,std,
                 age_group_bounds = boundaries,
                 detail=TRUE)
    
    show(this_fit, hue='blue')
}
```

# Example: Super-small populations with Italy 1980 rates

Repeat the exercise in the last section but with simulated deaths for a much smaller population subject to the same rates.  Suppose there were only 1000 Italian women and simulate deaths:

```{r}


for (i in 1:4) {  
  ## simulate a small population    
    target_pop = 1000
    smallN = N * target_pop/sum(N)
    smallD = rpois(length(smallN), smallN* D/N)
    
    this_fit = TOPALS_fit(smallN, smallD,std,
                 age_group_bounds = boundaries,
                 detail=TRUE)
    
    show(this_fit, hue='darkgreen')
}

```
